A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation

### 2023-06-05 paper/paper1/ms.tex
CVM rewrites the title! Much better.
Start of major re-arrangement of Sections 4 and 5, per CVM draft comments

\title{A large leading tail of atmospheric escape in the HAT-P-67 b system}

@@

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis} we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in \ref{secHeAnalysis}, with deep dives on the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with both 1D and 3D models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysicalMechanisms}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

\section{Causal Interpretations} \label{secPhysicalMechanisms}

\subsection{Causes for the leading tail} \label{secLeading}
Several physical phenomena could give rise to a leading tail geometry.  Here we explore the predominant two physical effects: orbital shear and stellar wind confinement.

@@

\subsection{Ohmic Dissipation and runaway escape}
\object{HAT-P-67 b} exhibits both dramatic atmospheric escape and significant radius inflation.  Ohmic dissipation stands out as a physical mechanism that could be responsible for both of these outstanding attributes.  \citet{2011ApJ...738....1B} showed that Ohmic heating acts to inflate hot Jupiters, with planets $\lesssim 0.5 M_\mathrm{Jup}$ overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The Ohmic heating scenario requires only a modest planetary magnetic field ($\gtrsim$1 G) and an equilibrium temperature great enough to thermally ionize some modest fraction of neutral metals, such as the low ionization species of \ion{Na}{1} and \ion{K}{1}.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K \citep{2011ApJ...738....1B,2012ApJ...745..138M,2016ApJ...819..116G,2018AJ....155..214T,2022A&A...658L...7K}, where the conductivity is strong enough to cause an effective drag without being so strong that magnetic braking slows the planetary wind.  HAT-P-67 b's $\sim$2000 K sits to the higher end, but still in a region of high Ohmic dissipation heating efficiency.  A proposed order-of-magnitude scaling law predicts an inflation timescale, Eq. 20 in \citet{2011ApJ...738....1B}:

A few caveats complicate the unambiguous causal interpretation of Ohmic dissipation.  Chiefly, numerical simulations \citep{2013ApJ...763...13W} and analytic theory \citep{2016ApJ...819..116G} indicate that Ohmic dissipation can stall the contraction of hot Jupiters, but cannot easily ``re-inflate'' them after having undergone a traditional cooling curve.  Heat transfer from the atmosphere to a cooled core simply proceeds too slowly, on the order of tens of Gyr \citep{2016ApJ...819..116G}.  This path dependence of Ohmic dissipation would restrict the allowed evolutionary histories of \hatpb, to have arrived at its current location within a few to tens of Myr.  This short time window prefers a physical mechanism such as disk migration, which would be faster than secular eccentric migration with an outer companion. \emph{In-situ} formation of a hot Saturn at these close-in separations may be implausible \citep{2018ARA&A..56..175D}.  Other caveats like unknowns in planetary magnetic fields, zonal band geometries, and dayside/nightside temperature differences make it impossible to uniquely prescribe Ohmic dissipation, and instead, several additional factors may also be at play \citep{2021A&A...645A..79S}.

\subsection{XUV Irradiation-driven mass loss}
XUV photoevaporation heats the upper layers in the atmosphere, driving a hydrodynamic wind \citep{2009ApJ...693...23M}.  Photoevaporation stands out as offering a natural cause of the observed leading tail attributable to dayside/nightside differences in mass loss: the greatest supersonic motions arise from the sub-solar point on the planetary dayside.  The effect may be boosted if dayside/nightside energy transport proceeds inefficiently \citep{2009ApJ...693...23M}.  Such a scenario is illustrated in Figure \ref{fig:KeplerianShear}.

@@

\subsection{Constraining mass loss rate}
The theory-inspired mass loss rates above correspond to 1D Parker wind \ion{He}{1} feature depths and widths similar to what we observe, as evinced in Figure \ref{fig:pwinds}, with $\dot{M}_\mathrm{1D}\sim100$ M$_\oplus$/Gyr.

First, the scaling law arguments that produced Equation \ref{eqInflate} were only proposed as order of magnitude estimates not to be taken too faithfully.  So a $10\times$ higher inflation timescale of 50 Myr would yield $\dot{M}_\mathrm{infl}\sim10^{14}$ g/s, still a very large mass loss rate, but only a factor of 5 away from the baseline 1D model.  Factors of a few uncertainties in the Ohmic dissipation efficiency $\epsilon$ may also contribute.

As mentioned in Section \ref{pwinds}, inferring the mass loss rate from observations is strongly model dependent.  The ionization fraction profile, exosphere temperature $T_0$, and $\dot{M}$ are all at least partially degenerate.  In particular, the uncertain XUV-to-NUV flux ratio confounds the interpretation of the observed Helium signal.  Even the ``high XUV'' scenario in Figure \ref{fig:XUV} does not have enough ground-state ionizing photons to penetrate to within 5 $R_p$ of the planet's surface, in a 1D Parker wind model.  This means that at best only 1 in $10^{6}$ Helium atoms participate in the observability of the Helium metastable triplet lines in the baseline 1D Parker wind model.  Turning down the XUV flux makes the signal even less observable.  A larger mass loss rate would be needed to explain the same observed HPF Helium absorption, if say only 1 in $10^{7}$ helium atoms are populating the metastable state.

>>>>

\title{A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation}

@@

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis} we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in \ref{secHeAnalysis}, with deep dives on the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with both 1D and 3D models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

\subsection{Causes for the leading tail} \label{secLeading}
Several physical phenomena could give rise to a leading tail geometry.  Here we explore the predominant two dynamical effects: orbital shear and stellar wind confinement.

@@

\section{Mechanisms Driving Atmospheric Escape}\label{secPhysMech}

\subsection{XUV Irradiation-driven runaway}
XUV photoevaporation heats the upper layers in the atmosphere, driving a hydrodynamic wind \citep{2009ApJ...693...23M}.  Photoevaporation stands out as offering a natural cause of the observed leading tail attributable to dayside/nightside differences in mass loss: the greatest supersonic motions arise from the sub-solar point on the planetary dayside.  The effect may be boosted if dayside/nightside energy transport proceeds inefficiently \citep{2009ApJ...693...23M}.  Such a scenario is illustrated in Figure \ref{fig:KeplerianShear}.

@@

\subsection{Ohmic Dissipation-driven runaway}
Ohmic dissipation stands out as a physical mechanism that could be responsible for both dramatic atmospheric escape and its significant radius inflation.  \citet{2011ApJ...738....1B} showed that Ohmic heating acts to inflate hot Jupiters, with planets $\lesssim 0.5 M_\mathrm{Jup}$ overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The Ohmic heating scenario requires only a modest planetary magnetic field ($\gtrsim$1 G) and an equilibrium temperature great enough to thermally ionize some modest fraction of neutral metals, such as the low ionization species of \ion{Na}{1} and \ion{K}{1}.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K \citep{2011ApJ...738....1B,2012ApJ...745..138M,2016ApJ...819..116G,2018AJ....155..214T,2022A&A...658L...7K}, where the conductivity is strong enough to cause an effective drag without being so strong that magnetic braking slows the planetary wind.  HAT-P-67 b's $\sim$2000 K sits to the higher end, but still in a region of high Ohmic dissipation heating efficiency.  A proposed order-of-magnitude scaling law predicts an inflation timescale, Eq. 20 in \citet{2011ApJ...738....1B}:

A few caveats complicate the unambiguous causal interpretation of Ohmic dissipation.  First, the scaling law arguments that produced Equation \ref{eqInflate} were only proposed as coarse scaling law estimates, with numerical simulations needed to quantify inflation timescales for individual systems.  Accordingly, a $10\times$ higher inflation timescale of 50 Myr---allowed by the coarse scaling law---would yield $\dot{M}_\mathrm{infl}\sim10^{14}$ g/s, still a very large mass loss rate, and only a factor of 5 away from the baseline 1D model.  Factors of a few uncertainties in the Ohmic dissipation efficiency $\epsilon$ may also contribute.

Second, numerical simulations \citep{2013ApJ...763...13W} and analytic theory \citep{2016ApJ...819..116G} indicate that Ohmic dissipation can stall the contraction of hot Jupiters, but cannot easily ``re-inflate'' them after having undergone a traditional cooling curve.  Heat transfer from the atmosphere to a cooled core appears to proceed too slowly, on the order of tens of Gyr \citep{2016ApJ...819..116G}.  This path dependence of Ohmic dissipation would restrict the allowed evolutionary histories of \hatpb, to have arrived at its current location within a few to tens of Myr.  This short time window prefers a physical mechanism such as disk migration, which would be faster than secular eccentric migration with an outer companion. \emph{In-situ} formation of a hot Saturn at these close-in separations may be implausible \citep{2018ARA&A..56..175D}.

Other caveats like unknowns in planetary magnetic fields, zonal band geometries, and dayside/nightside temperature differences make it impossible to uniquely prescribe Ohmic dissipation, and instead, several additional factors may also be at play \citep{2021A&A...645A..79S}.
